---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# stevents

<!-- badges: start -->
[![R-CMD-check](https://github.com/ricardosp4/stevents/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ricardosp4/stevents/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

> Lightweight S3 classes for spatio-temporal event pattern analysis in R

`stevents` provides tools for representing, manipulating, and analysing
spatio-temporal point patterns. It was designed with wildfire ignition
modelling in mind, but applies to any domain where events have both a
location and a timestamp — seismology, epidemiology, criminology.

## Installation

```r
devtools::install_github("ricardosp4/stevents")
```

## Core classes

| Class | Description |
|---|---|
| `stevents` | A set of events with coordinates and timestamps |
| `stgrid` | A regular space-time tessellation for aggregating events |

Each class has `print()`, `summary()`, and `plot()` methods.

## Functions

| Function | Description |
|---|---|
| `count_events()` | Count events per grid cell and time bin |
| `neighbor_history()` | Count recent nearby events for each event (discrete excitation) |
| `hawkes_intensity()` | Compute the Hawkes self-excitation kernel (continuous) |
| `as_sf()` | Convert `stevents` to an `sf` point layer |
| `subset()` | Filter events by time window and/or bounding box |

## Quick start

```{r quickstart, message = FALSE}
library(stevents)

set.seed(1)
ev <- stevents(
  x    = runif(200, 0, 100),
  y    = runif(200, 0, 100),
  time = as.POSIXct("2023-01-01") + sort(runif(200, 0, 365 * 86400))
)

print(ev)
summary(ev)
```

```{r plot-ev, fig.height = 4}
plot(ev)
```

## Hawkes kernel

```{r hawkes, message = FALSE}
ev2 <- hawkes_intensity(ev,
  alpha = 0.8,
  beta  = 1 / (7 * 86400),
  gamma = 1 / 30
)

summary(ev2$data$hawkes_intensity)
```

The continuous kernel captures finer variation than a fixed binary
window — events outside a discrete neighbourhood can still carry
non-trivial excitation.

```{r compare, fig.height = 4}
ev3 <- neighbor_history(ev, radius = 15, lag = 30 * 86400)
ev3 <- hawkes_intensity(ev3, alpha = 0.8,
                        beta  = 1 / (7 * 86400),
                        gamma = 1 / 30)

plot(
  ev3$data$n_recent_neighbors,
  ev3$data$hawkes_intensity,
  xlab = "n_recent_neighbors (discrete)",
  ylab = "Hawkes intensity (continuous)",
  pch  = 16,
  col  = adjustcolor("#E05A2B", 0.4),
  main = "Discrete vs continuous excitation"
)
```

## Included datasets

| Dataset | Description | Source |
|---|---|---|
| `venezuela_2026` | Seismic sequence following the Mw 7.5 earthquake of 24 June 2026 | USGS |
| `amatrice_2016` | Amatrice–Norcia aftershock sequence, August–November 2016 (74 events) | USGS |

```{r datasets, message = FALSE}
data(amatrice_2016)

events_italy <- stevents(
  x    = amatrice_2016$lon,
  y    = amatrice_2016$lat,
  time = amatrice_2016$time
)

print(events_italy)
```

## Motivation

`stevents` was developed as part of a Master's thesis on
spatio-temporal Hawkes process modelling of wildfire ignition risk on
the Iberian Peninsula. The package provides the preprocessing layer
between raw event data and intensity model fitting via INLA.

## License

MIT © Ricardo
