---
title: "Introduction to stevents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to stevents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Overview

`stevents` provides two S3 classes for spatio-temporal event analysis:

- `stevents`: a set of events with coordinates and timestamps. In addition
  to `print()`, `summary()`, and `plot()`, it has two extra methods:
  `subset()` (filter by time/space) and `as_sf()` (convert to `sf`).
- `stgrid`: a regular space-time tessellation for aggregating events. In
  addition to `print()`, `summary()`, and `plot()`, it has two extra
  methods: `dim()` (data-cube size) and `as_sf()` (convert cells to `sf`).

Three functions connect them:

- `count_events()`: counts events per grid cell and time bin.
- `neighbor_history()`: counts recent nearby events for each event.
- `hawkes_intensity()`: computes the Hawkes self-excitation kernel.

## Creating an `stevents` object

An `stevents` object requires x-coordinates, y-coordinates, and
timestamps. Events are sorted by time on construction.

```{r create-stevents}
library(stevents)

set.seed(1)
n <- 200

ev <- stevents(
  x    = runif(n, 0, 100),
  y    = runif(n, 0, 100),
  time = as.POSIXct("2023-01-01") + sort(runif(n, 0, 365 * 86400))
)
```

`print()` shows a compact summary of the object:

```{r print-stevents}
print(ev)
```

`summary()` reports derived statistics:

```{r summary-stevents}
summary(ev)
```

`plot()` shows the spatial distribution and temporal intensity:

```{r plot-stevents}
plot(ev)
```

`subset()` filters by time window, bounding box, or both:

```{r subset-stevents}
ev_q1 <- subset(ev,
  t_from = as.POSIXct("2023-01-01"),
  t_to   = as.POSIXct("2023-03-31")
)
print(ev_q1)
```

`as_sf()` converts the object to an `sf` point layer for use with
other spatial packages:

```{r as-sf}
ev_sf <- as_sf(ev)
class(ev_sf)
```

## Creating an `stgrid` object

An `stgrid` defines a regular tessellation over a bounding box and
time window.

```{r create-stgrid}
g <- stgrid(
  bbox      = c(0, 0, 100, 100),
  cell_size = 20,
  t_window  = as.POSIXct(c("2023-01-01", "2023-12-31")),
  t_step    = "1 month"
)

print(g)
summary(g)
```

`plot()` draws the spatial cells:

```{r plot-stgrid}
plot(g)
```

`dim()` reports the size of the underlying space-time data cube
(spatial cells x temporal bins):

```{r dim-stgrid}
dim(g)
```

`as_sf()` extracts the cells as a standalone `sf` polygon layer, for use
with the wider `sf`/`ggplot2` ecosystem:

```{r as-sf-stgrid}
sf_g <- as_sf(g)
class(sf_g)
head(sf_g)
```

## Counting events

`count_events()` assigns each event to a cell and time bin:

```{r count-events}
g2 <- count_events(ev, g)
head(g2$counts)
```

Cells are shaded by event count when plotted:

```{r plot-counts}
plot(g2)
```

## Neighbour history

`neighbor_history()` counts how many previous events occurred within
a given spatial radius and time lag for each event. This is a discrete
approximation of the self-excitation signal in a Hawkes process.

```{r neighbor-history}
ev2 <- neighbor_history(ev, radius = 15, lag = 30 * 86400)
summary(ev2$data$n_recent_neighbors)
```

## Hawkes intensity

`hawkes_intensity()` computes a continuous alternative using an
exponential decay kernel:

$$\Lambda_i = \sum_{j:\, t_j < t_i} \alpha \cdot e^{-\beta(t_i - t_j)} \cdot e^{-\gamma \, d_{ij}}$$

where $\alpha$ controls excitation magnitude, $\beta$ controls temporal
decay, and $\gamma$ controls spatial decay.

```{r hawkes-intensity}
ev3 <- hawkes_intensity(
  ev2,
  alpha = 0.5,
  beta  = 1 / (7 * 86400),
  gamma = 1 / 15
)
summary(ev3$data$hawkes_intensity)
```

Events with zero discrete neighbours can carry non-zero Hawkes
intensity from events just outside the fixed window:

```{r compare}
plot(
  ev3$data$n_recent_neighbors,
  ev3$data$hawkes_intensity,
  xlab = "n_recent_neighbors (discrete)",
  ylab = "Hawkes intensity (continuous)",
  pch  = 16,
  col  = adjustcolor("#E05A2B", 0.4)
)
```

## LLM usage

The LLM tool used was Claude (Sonnet), for writing and debugging R code
for the package functions and S3 methods, as well as drafting and
editing the text of this vignette and the "Introduction to stevents"
vignette. Approximately 60-70% of the code was generated by the LLM,
with all LLM-assisted code reviewed, tested, and adjusted by the author
before inclusion in the package.
