---
title: "Spatio-temporal Hawkes processes: real seismic sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal Hawkes processes: real seismic sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Overview

Seismic sequences are a canonical application of Hawkes process models.
Each earthquake raises the probability of subsequent earthquakes through
stress transfer — a mechanism described by the Omori–Utsu law, which is
a special case of the exponential decay kernel implemented in
`hawkes_intensity()`.

This vignette applies `stevents` to two real seismic sequences downloaded
directly from the USGS Earthquake Catalog API:

- **Venezuela 2026**: a Mw 7.5 mainshock on 24 June 2026, one of the
  strongest earthquakes in Venezuela in over a century. The sequence is
  recent and the aftershock catalog is still growing.
- **Amatrice 2016**: a Mw 6.2 mainshock on 24 August 2016 in central
  Italy, followed by a Mw 6.5 event on 30 October. With 76 recorded
  aftershocks over three months, this sequence provides a complete view
  of aftershock decay and self-excitation.

Together, the two cases illustrate how `stevents` handles real
spatio-temporal point patterns at different stages of completeness.

## Case study 1: Venezuela 2026

On 24 June 2026, two major earthquakes struck northern Venezuela within
39 seconds: a Mw 7.2 foreshock followed by a Mw 7.5 mainshock, both
near Yumare in the state of Yaracuy. The sequence caused at least 920
deaths and more than 4,500 injuries.

Seismic sequences are a natural application of Hawkes process models.
Each earthquake raises the probability of subsequent earthquakes through
stress transfer, a mechanism known as Omori–Utsu aftershock decay, which
is a special case of the exponential kernel implemented in `stevents`.

This vignette downloads the complete aftershock sequence from the USGS
Earthquake Catalog API and analyses it using `stevents`.

## Data

The USGS FDSN API provides programmatic access to earthquake catalogs.
We query all events of magnitude 2.5 or greater within 200 km of the
mainshock epicenter between 24 June and 27 June 2026.

```{r download, message = FALSE}
library(stevents)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

url <- paste0(
  "https://earthquake.usgs.gov/fdsnws/event/1/query?format=csv",
  "&starttime=2026-06-22",
  "&endtime=2026-07-27",
  "&minmagnitude=2.5",
  "&latitude=10.435",
  "&longitude=-68.472",
  "&maxradiuskm=200",
  "&orderby=time"
)

raw <- read.csv(url(url), stringsAsFactors = FALSE)
```

We retain only the columns needed and convert timestamps to `POSIXct`:

```{r clean}
eq <- data.frame(
  lon       = raw$longitude,
  lat       = raw$latitude,
  depth_km  = raw$depth,
  magnitude = raw$mag,
  time      = as.POSIXct(raw$time, format = "%Y-%m-%dT%H:%M:%OS", tz = "UTC")
)

nrow(eq)
range(eq$magnitude)
range(eq$time)
```

## Creating an `stevents` object

Coordinates are in geographic degrees (WGS84). We project to UTM Zone
19N (EPSG:32619), the metric CRS appropriate for northern Venezuela, so
that distances in `neighbor_history()` and `hawkes_intensity()` are in
metres.

```{r project}
pts <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)
pts_utm <- st_transform(pts, crs = 32619)
coords  <- st_coordinates(pts_utm)

events <- stevents(
  x    = coords[, 1],
  y    = coords[, 2],
  time = eq$time,
  crs  = 32619
)

print(events)
```

## Spatial distribution

```{r map, fig.height = 5, fig.width = 7, message = FALSE, warning = FALSE}
venezuela <- ne_countries(
  country     = "Venezuela",
  scale       = "medium",
  returnclass = "sf"
)

eq_sf <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)

bbox_v  <- st_bbox(eq_sf)
buffer  <- 0.5  # degrees

ggplot() +
  geom_sf(data = venezuela, fill = "#F5F0E8", colour = "grey60",
          linewidth = 0.3) +
  geom_sf(data   = eq_sf,
          aes(colour = magnitude),
          size   = 2.5,
          alpha  = 0.8) +
  scale_colour_gradient(low = "#FD8D3C", high = "#8B2500",
                        name = "Magnitude") +
  coord_sf(
    xlim = c(bbox_v["xmin"] - buffer, bbox_v["xmax"] + buffer),
    ylim = c(bbox_v["ymin"] - buffer, bbox_v["ymax"] + buffer)
  ) +
  labs(
    title   = "Seismic sequence. Venezuela, 24–27 June 2026",
    caption = "Source: USGS Earthquake Catalog"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 7, colour = "grey60")
  )
```

## Temporal pattern

```{r plot-events, fig.height = 4}
plot(events)
```

The temporal plot shows the characteristic aftershock decay: high
activity immediately after the mainshock, declining rapidly over hours
and days. This pattern is described by the Omori–Utsu law.

## Neighbour history

We compute the discrete self-excitation signal using a 50 km radius and
a 6-hour lag, parameters appropriate for aftershock triggering at this
spatial and temporal scale.

```{r neighbor}
events_hist <- neighbor_history(
  events,
  radius = 50000,      # 50 km
  lag    = 6 * 3600    # 6 hours
)

summary(events_hist$data$n_recent_neighbors)
```

## Hawkes intensity

The continuous kernel uses the same spatial scale but replaces the
binary window with exponential decay. We set the temporal decay to
approximately 2 hours, consistent with early aftershock sequences.

```{r hawkes}
events_hawkes <- hawkes_intensity(
  events_hist,
  alpha = 0.8,
  beta  = 1 / (2 * 3600),   # decay scale: ~2 hours
  gamma = 1 / 50000          # decay scale: ~50 km
)

summary(events_hawkes$data$hawkes_intensity)
```

The intensity peaks immediately after the mainshock and decays as the
sequence quiets:

```{r hawkes-plot, fig.height = 4}
plot(
  events_hawkes$data$time,
  events_hawkes$data$hawkes_intensity,
  type = "l",
  col  = "#E05A2B",
  xlab = "Time (UTC)",
  ylab = "Hawkes intensity",
  main = "Conditional intensity. Venezuela aftershock sequence"
)
```

## Discrete vs continuous excitation

```{r compare, fig.height = 4}
plot(
  events_hawkes$data$n_recent_neighbors,
  events_hawkes$data$hawkes_intensity,
  xlab = "n_recent_neighbors (discrete)",
  ylab = "Hawkes intensity (continuous)",
  pch  = 16,
  col  = adjustcolor("#E05A2B", 0.5),
  main = "Discrete vs continuous excitation"
)
abline(lm(hawkes_intensity ~ n_recent_neighbors,
          data = events_hawkes$data),
       col = "#8B2500", lwd = 2, lty = 2)
```

## Case study 2: Amatrice, Italy 2016

The Amatrice sequence is one of the most studied seismic sequences in
recent European history. A Mw 6.2 mainshock struck central Italy on 24
August 2016, followed by a second major event of Mw 6.5 on 30 October
2016, stronger than the first, which reactivated the sequence and
produced a new wave of aftershocks. The full three-month catalog
contains 76 events of magnitude 2.5 or greater, making it a richer
dataset for illustrating self-excitation than a simple
mainshock–aftershock sequence.

```{r italy-download, message = FALSE}
url_italy <- paste0(
  "https://earthquake.usgs.gov/fdsnws/event/1/query?format=csv",
  "&starttime=2016-08-20",
  "&endtime=2016-12-20",
  "&minmagnitude=2.5",
  "&latitude=42.698",
  "&longitude=13.234",
  "&maxradiuskm=100",
  "&orderby=time"
)

raw_italy <- read.csv(url(url_italy), stringsAsFactors = FALSE)

italy <- data.frame(
  lon       = raw_italy$longitude,
  lat       = raw_italy$latitude,
  depth_km  = raw_italy$depth,
  magnitude = raw_italy$mag,
  time      = as.POSIXct(raw_italy$time,
                          format = "%Y-%m-%dT%H:%M:%OS",
                          tz     = "UTC")
)
```

```{r italy-project}
pts_italy     <- st_as_sf(italy, coords = c("lon", "lat"), crs = 4326)
pts_italy_utm <- st_transform(pts_italy, crs = 32633)  # UTM zone 33N for Italy
coords_italy  <- st_coordinates(pts_italy_utm)

events_italy <- stevents(
  x    = coords_italy[, 1],
  y    = coords_italy[, 2],
  time = italy$time,
  crs  = 32633
)

print(events_italy)
```

```{r italy-map, fig.height = 5, fig.width = 7, message = FALSE, warning = FALSE}
italy_map <- ne_countries(
  country     = "Italy",
  scale       = "medium",
  returnclass = "sf"
)

bbox_i <- st_bbox(pts_italy)
buffer <- 0.15  # degrees

ggplot() +
  geom_sf(data = italy_map, fill = "#F5F0E8", colour = "grey60",
          linewidth = 0.3) +
  geom_sf(data   = pts_italy,
          aes(colour = magnitude),
          size   = 2.5,
          alpha  = 0.8) +
  scale_colour_gradient(low = "#FD8D3C", high = "#8B2500",
                        name = "Magnitude") +
  coord_sf(
    xlim = c(bbox_i["xmin"] - buffer, bbox_i["xmax"] + buffer),
    ylim = c(bbox_i["ymin"] - buffer, bbox_i["ymax"] + buffer)
  ) +
  labs(
    title   = "Amatrice aftershock sequence. August–November 2016",
    caption = "Source: USGS Earthquake Catalog"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 7, colour = "grey60")
  )
```

```{r italy-plot}
plot(events_italy)
```

The temporal histogram shows the characteristic aftershock decay: high
activity in the first days after the mainshock, declining progressively
over the following weeks.

```{r italy-hawkes}
events_italy_h <- neighbor_history(
  events_italy,
  radius = 30000,     # 30 km
  lag    = 3 * 86400  # 3 days
)

events_italy_h <- hawkes_intensity(
  events_italy_h,
  alpha = 0.8,
  beta  = 1 / (2 * 86400),  # decay scale: ~2 days
  gamma = 1 / 30000          # decay scale: ~30 km
)

summary(events_italy_h$data$hawkes_intensity)
```

```{r italy-intensity-plot, fig.height = 4}
plot(
  events_italy_h$data$time,
  events_italy_h$data$hawkes_intensity,
  type = "l",
  col  = "#E05A2B",
  xlab = "Time (UTC)",
  ylab = "Hawkes intensity",
  main = "Conditional intensity. Amatrice sequence"
)
```

## Summary

This vignette applied `stevents` to two real seismic sequences:

- **Venezuela 2026**: an ongoing sequence captured in near real time,
  demonstrating the full `stevents` workflow with live USGS data.
- **Amatrice 2016**: a complete aftershock sequence with 76 recorded
  events, showing the Omori–Utsu temporal decay and a clear
  self-excitation signal in the Hawkes intensity.

Both sequences illustrate the same principle: past seismic activity
raises the conditional probability of future events, and
`hawkes_intensity()` quantifies this effect continuously in space
and time. Estimating the kernel parameters from data via maximum
likelihood is the natural next step toward a fully calibrated
spatio-temporal Hawkes model.
