---
title: "Spatio-temporal Hawkes processes: real seismic sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal Hawkes processes: real seismic sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Overview

Seismic sequences are a canonical application of Hawkes process models:
each earthquake raises the probability of subsequent earthquakes through
stress transfer, a mechanism described by the Omori-Utsu law and
captured here by the exponential kernel in `hawkes_intensity()`. This
vignette applies `stevents` to two real earthquake catalogs downloaded
from the USGS Earthquake Catalog API, and uses them together to
illustrate the package's workflow and the difference between a
*complete* and a *still-growing* catalog.

```{r libs, message = FALSE}
library(stevents)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Data

- **Amatrice 2016** (Italy): a Mw 6.2 mainshock on 24 August 2016,
  followed by a Mw 6.5 event on 30 October 2016. The bundled catalog
  covers three months and 74 events of magnitude 2.5+, i.e. a
  post-hoc, essentially complete record of the sequence.
- **Venezuela 2026**: a Mw 7.5 mainshock on 24 June 2026 near Yumare.
  This catalog is downloaded from the mainshock up to the download
  date (see `data-raw/venezuela_2026.R`), so its size grows every time
  it is refreshed, as USGS reviews and adds aftershocks. This is real,
  unmodified API output - it illustrates a common situation in
  operational seismology, where a catalog for a recent sequence is
  incomplete until enough time has passed for full review.

```{r clean}
data(amatrice_2016)
data(venezuela_2026)
nrow(amatrice_2016); range(amatrice_2016$magnitude)
nrow(venezuela_2026); range(venezuela_2026$magnitude)
```

Because the two catalogs differ in completeness, Amatrice is used
below for the full workflow (temporal pattern, neighbour counts,
Hawkes intensity). For Venezuela, the same self-excitation analysis is
only run if the catalog is large enough (arbitrarily, at least 20
events) to give a meaningful signal; otherwise a handful of events
would produce a noisy, misleading estimate, so the vignette falls back
to a map and summary only.

```{r venezuela-threshold}
ven_min_n     <- 20
ven_full_run  <- nrow(venezuela_2026) >= ven_min_n
ven_full_run
```

## Spatial distribution

Both sequences are mapped side by side, coloured by magnitude, each on
its own country outline for geographic context.

```{r maps, fig.height = 4.5, fig.width = 8, message = FALSE, warning = FALSE}
make_map <- function(eq, country, label) {
  bg  <- ne_countries(country = country, scale = "medium", returnclass = "sf")
  pts <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)
  bb  <- st_bbox(pts)
  buf <- 0.3

  ggplot() +
    geom_sf(data = bg, fill = "#F5F0E8", colour = "grey60", linewidth = 0.3) +
    geom_sf(data = pts, aes(colour = magnitude), size = 2.5, alpha = 0.85) +
    scale_colour_gradient(low = "#FD8D3C", high = "#8B2500", name = "Mag.") +
    coord_sf(xlim = c(bb["xmin"] - buf, bb["xmax"] + buf),
             ylim = c(bb["ymin"] - buf, bb["ymax"] + buf)) +
    labs(title = paste0(label, " (n = ", nrow(eq), ")")) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(size = 11, face = "bold"))
}

gridExtra_ok <- requireNamespace("gridExtra", quietly = TRUE)
p1 <- make_map(amatrice_2016,  "Italy",     "Amatrice 2016")
p2 <- make_map(venezuela_2026, "Venezuela", "Venezuela 2026")

if (gridExtra_ok) {
  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  p1
  p2
}
```

The Amatrice map shows a dense, elongated cloud of aftershocks tracing
the ruptured fault segment. The Venezuela map shows the mainshock and
its early aftershocks near Yumare, still a much sparser cloud than
Amatrice's - consistent with a catalog still being populated.

## Building `stevents` objects

Coordinates are projected to a local metric CRS (UTM) so that distances
used later by `neighbor_history()` and `hawkes_intensity()` are in
metres.

```{r build}
to_stevents <- function(eq, epsg) {
  pts  <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)
  utm  <- st_transform(pts, crs = epsg)
  xy   <- st_coordinates(utm)
  stevents(x = xy[, 1], y = xy[, 2], time = eq$time, crs = epsg)
}

events_italy <- to_stevents(amatrice_2016,  32633)  # UTM 33N
events_ven   <- to_stevents(venezuela_2026, 32619)  # UTM 19N

print(events_italy)
print(events_ven)
```

## Temporal pattern (Amatrice)

`plot()` draws the spatial scatter and a temporal histogram side by
side. With 74 events spread over three months, the histogram is
informative; the much smaller Venezuela catalog (`r nrow(venezuela_2026)`
events so far) would show only a handful of bars, so it is omitted here
and revisited below.

```{r plot-events}
plot(events_italy)
```

The left panel shows where events occurred (an elongated cloud along
the fault); the right panel shows event counts over time. Two spikes
are visible: the initial burst of aftershocks right after the 24
August mainshock, and a second, larger burst after the 30 October
Mw 6.5 event that reactivated the sequence - the classic Omori-Utsu
decay pattern, restarted by a secondary event.

## Self-excitation: neighbour history and Hawkes intensity

`neighbor_history()` counts, for each event, how many earlier events
occurred within a spatial radius and time lag - a discrete measure of
local recent activity. `hawkes_intensity()` computes a continuous
alternative using an exponential decay kernel in both time and space.

```{r hawkes}
events_h <- neighbor_history(events_italy, radius = 30000, lag = 3 * 86400)
events_h <- hawkes_intensity(
  events_h,
  alpha = 0.8,
  beta  = 1 / (2 * 86400),  # ~2-day temporal decay
  gamma = 1 / 30000          # ~30 km spatial decay
)
summary(events_h$data$hawkes_intensity)
```

```{r hawkes-plot}
plot(
  events_h$data$time, events_h$data$hawkes_intensity,
  type = "l", col = "#E05A2B",
  xlab = "Time (UTC)", ylab = "Hawkes intensity",
  main = "Conditional intensity - Amatrice sequence"
)
```

This is the **conditional intensity** implied by the Hawkes kernel: how
"excited" the process is at each moment, given all earlier events nearby
in space and time. It spikes sharply after each mainshock and decays
smoothly between events, unlike the discrete, blocky counts in
`neighbor_history()`.

```{r compare}
plot(
  events_h$data$n_recent_neighbors, events_h$data$hawkes_intensity,
  xlab = "n_recent_neighbors (discrete)",
  ylab = "Hawkes intensity (continuous)",
  pch  = 16, col = adjustcolor("#E05A2B", 0.5),
  main = "Discrete vs continuous excitation"
)
abline(lm(hawkes_intensity ~ n_recent_neighbors, data = events_h$data),
       col = "#8B2500", lwd = 2, lty = 2)
```

Each point compares, for one event, its discrete neighbour count
against its continuous Hawkes intensity. The positive trend (dashed
line) confirms the two measures agree on average, but the scatter shows
they are not interchangeable: `hawkes_intensity()` also credits events
just outside the fixed radius/lag window, weighted by distance and
recency, so two events with the same discrete count can have different
continuous intensities.

## Venezuela: a growing catalog

Whether we go further than the map and summary above depends on
`ven_full_run` (defined earlier): with fewer than `r ven_min_n` events,
`neighbor_history()` and `hawkes_intensity()` would run without error,
but their output would not be a meaningful estimate of self-excitation
- a realistic constraint of working with live catalogs, not a
limitation of the package. When the catalog has grown enough, the same
two-step workflow used for Amatrice is applied here as well.

```{r venezuela-full, eval = ven_full_run}
events_ven_h <- neighbor_history(events_ven, radius = 30000, lag = 3 * 86400)
events_ven_h <- hawkes_intensity(
  events_ven_h,
  alpha = 0.8, beta = 1 / (2 * 86400), gamma = 1 / 30000
)
summary(events_ven_h$data$hawkes_intensity)
plot(
  events_ven_h$data$time, events_ven_h$data$hawkes_intensity,
  type = "l", col = "#E05A2B",
  xlab = "Time (UTC)", ylab = "Hawkes intensity",
  main = "Conditional intensity - Venezuela sequence"
)
```

```{r venezuela-message, eval = !ven_full_run, echo = FALSE, results = "asis"}
cat(
  "*(Skipped: the current Venezuela catalog has", nrow(venezuela_2026),
  "events, below the", ven_min_n, "-event threshold set above.)*"
)
```

## Discussion

Applying the same functions to a complete catalog (Amatrice) and a
still-growing one (Venezuela) shows both the strength and a limit of
the workflow: `stevents` and `stgrid` handle real, irregular
spatio-temporal point patterns directly from `sf`/`POSIXct` data, and
`hawkes_intensity()` recovers a smooth, physically interpretable
excitation signal from a well-sampled sequence. But the method is only
as good as the data feeding it - a useful reminder when applying it to
real-time monitoring rather than retrospective analysis. Estimating the
kernel parameters from data via maximum likelihood, rather than setting
them by hand as done here, is the natural next step.

## LLM usage

- **LLM tool used**: Claude (Sonnet).
- **Purpose**: writing and debugging R code for the package functions and
  S3 methods, and drafting/editing the text of this vignette and the
  "Introduction to stevents" vignette.
- **Percentage of code generated by the LLM**: approximately 50-75%.
  All LLM-assisted code was reviewed, tested, and adjusted by the author
  before inclusion in the package.
