---
title: "Spatio-temporal Hawkes processes: real seismic sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal Hawkes processes: real seismic sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Overview

Seismic sequences are a canonical application of Hawkes process models:
each earthquake raises the probability of subsequent earthquakes through
stress transfer, a mechanism described by the Omori-Utsu law and
captured here by the exponential kernel in `hawkes_intensity()`. This
vignette applies `stevents` to two real earthquake catalogs downloaded
from the USGS Earthquake Catalog API, and uses them together to
illustrate the package's workflow and the difference between a
complete and a still-growing catalog.

```{r libs, message = FALSE}
library(stevents)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Data

**Amatrice 2016** (Italy): a Mw 6.2 mainshock on 24 August 2016,
followed by a Mw 6.5 event on 30 October 2016. The bundled catalog
covers three months and 74 events of magnitude 2.5+, i.e. a
post-hoc, essentially complete record of the sequence.

**Venezuela 2026**: a Mw 7.5 mainshock on 24 June 2026 near Yumare.
This catalog is downloaded from the mainshock up to the download
date (see `data-raw/venezuela_2026.R`), so its size grows every time
it is refreshed, as USGS reviews and adds aftershocks. This is real,
unmodified API output, illustrating a common situation in
operational seismology: a catalog for a recent sequence stays
incomplete until enough time has passed for full review.

```{r clean}
data(amatrice_2016)
data(venezuela_2026)
nrow(amatrice_2016); range(amatrice_2016$magnitude)
nrow(venezuela_2026); range(venezuela_2026$magnitude)
```

Because the two catalogs differ in completeness, Amatrice is used
below for the full workflow (temporal pattern, neighbour counts,
Hawkes intensity). For Venezuela, the same self-excitation analysis is
only run if the catalog is large enough (arbitrarily, at least 20
events) to give a meaningful signal; otherwise a handful of events
would produce a noisy, misleading estimate, so the vignette falls back
to a map and summary only.

```{r venezuela-threshold}
ven_min_n     <- 20
ven_full_run  <- nrow(venezuela_2026) >= ven_min_n
ven_full_run
```

## Spatial distribution

Both sequences are mapped side by side, coloured by magnitude, each on
its own country outline for geographic context.

```{r maps, fig.height = 4.5, fig.width = 8, message = FALSE, warning = FALSE}
make_map <- function(eq, country, label) {
  bg  <- ne_countries(country = country, scale = "medium", returnclass = "sf")
  pts <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)
  bb  <- st_bbox(pts)
  buf <- 0.15

  ggplot() +
    geom_sf(data = bg, fill = "#F5F0E8", colour = "grey60", linewidth = 0.3) +
    geom_sf(data = pts, aes(colour = magnitude), size = 2.5, alpha = 0.85) +
    scale_colour_gradient(low = "#FD8D3C", high = "#8B2500", name = "Mag.") +
    coord_sf(xlim = c(bb["xmin"] - buf, bb["xmax"] + buf),
             ylim = c(bb["ymin"] - buf, bb["ymax"] + buf)) +
    labs(title = paste0(label, " (n = ", nrow(eq), ")")) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(size = 11, face = "bold"))
}

gridExtra_ok <- requireNamespace("gridExtra", quietly = TRUE)
p1 <- make_map(amatrice_2016,  "Italy",     "Amatrice 2016")
p2 <- make_map(venezuela_2026, "Venezuela", "Venezuela 2026")

if (gridExtra_ok) {
  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  p1
  p2
}
```

The Amatrice map shows a dense, elongated cloud of aftershocks tracing
the ruptured fault segment. The Venezuela map shows the mainshock and
its early aftershocks near Yumare, a much sparser cloud than
Amatrice's, consistent with a catalog that is still being populated.

## Building `stevents` objects

Coordinates are projected to a local metric CRS (UTM) so that distances
used later by `neighbor_history()` and `hawkes_intensity()` are in
metres.

```{r build}
to_stevents <- function(eq, epsg) {
  pts  <- st_as_sf(eq, coords = c("lon", "lat"), crs = 4326)
  utm  <- st_transform(pts, crs = epsg)
  xy   <- st_coordinates(utm)
  stevents(x = xy[, 1], y = xy[, 2], time = eq$time, crs = epsg)
}

events_italy <- to_stevents(amatrice_2016,  32633)  # UTM 33N
events_ven   <- to_stevents(venezuela_2026, 32619)  # UTM 19N

print(events_italy)
print(events_ven)
```

## Temporal pattern (Amatrice)

`plot()` draws the spatial scatter and a temporal histogram side by
side. With 74 events spread over three months, the histogram is
informative; the much smaller Venezuela catalog (`r nrow(venezuela_2026)`
events so far) would show only a handful of bars, so it is omitted here
and revisited below.

```{r plot-events}
plot(events_italy)
```

The left panel shows where events occurred, an elongated cloud along
the fault, and the right panel shows event counts over time. Two
spikes are visible: the initial burst of aftershocks right after the
24 August mainshock, and a second, larger burst after the 30 October
Mw 6.5 event. Both bursts decay over the following days, which is the
classic Omori-Utsu pattern: aftershock rate falls off roughly as a
power law after a mainshock, and here it is visibly restarted by the
second, comparably large event rather than following a single smooth
decay for the whole three-month window.

## Self-excitation: neighbour history and Hawkes intensity

`neighbor_history()` counts, for each event, how many earlier events
occurred within a spatial radius and time lag, a discrete measure of
local recent activity. `hawkes_intensity()` computes a continuous
alternative using an exponential decay kernel in both time and space.

**A note on the parameters used below:** `alpha`, `beta`, and `gamma`
are set here by hand, not estimated from the data. This vignette
demonstrates the mechanics of the kernel, not a fitted model: `beta`
corresponds to an e-folding time of about two days, and `gamma` to an
e-folding distance of about 30 km, chosen to roughly match the 3-day,
30 km window used for `neighbor_history()` just above so the two
measures are directly comparable in the plots that follow. These are
physically plausible values for a shallow M6+ mainshock-aftershock
sequence, but they are illustrative, not fitted. A proper analysis
would estimate all three parameters from the event times and
locations themselves, typically via maximum likelihood; that
estimation step is intentionally out of scope for this package and is
the natural next stage of the thesis work this package supports (see
the Discussion below).

```{r hawkes}
events_h <- neighbor_history(events_italy, radius = 30000, lag = 3 * 86400)
events_h <- hawkes_intensity(
  events_h,
  alpha = 0.8,
  beta  = 1 / (2 * 86400),  # ~2-day temporal decay
  gamma = 1 / 30000          # ~30 km spatial decay
)
summary(events_h$data$hawkes_intensity)
```

The minimum intensity in this summary is exactly zero, which
corresponds to an event with no recent, nearby predecessors at all:
the kernel sum is empty, so only the background contributes. The
median, 2.82, and mean, 3.19, sit well below the maximum, 8.57, and
the mean exceeds the median, indicating a right-skewed distribution:
most events carry a modest intensity, while a smaller number sit
much closer to the peak, immediately after a cluster of nearby
predecessors. That spread, from a hard floor at zero to a peak nearly
three times the mean, is the whole point of the model: a homogeneous
Poisson process would assign every event the same rate, and it is
precisely this spread that motivates treating the sequence as
self-exciting rather than uniformly random.

```{r hawkes-plot}
plot(
  events_h$data$time, events_h$data$hawkes_intensity,
  type = "l", col = "#E05A2B",
  xlab = "Time (UTC)", ylab = "Hawkes intensity",
  main = "Conditional intensity: Amatrice sequence"
)
```

This is the conditional intensity implied by the Hawkes kernel: how
"excited" the process is at each moment, given all earlier events nearby
in space and time. It spikes sharply after each mainshock and decays
smoothly between events, unlike the discrete, blocky counts in
`neighbor_history()`. The two spikes line up with the same two bursts
seen in the temporal histogram above, the 24 August mainshock and the
30 October Mw 6.5 reactivation, and each decays back toward baseline
over roughly the two-day timescale set by `beta`, exactly as that
parameter choice implies it should.

```{r compare}
plot(
  events_h$data$n_recent_neighbors, events_h$data$hawkes_intensity,
  xlab = "n_recent_neighbors (discrete)",
  ylab = "Hawkes intensity (continuous)",
  pch  = 16, col = adjustcolor("#E05A2B", 0.5),
  main = "Discrete vs continuous excitation"
)
abline(lm(hawkes_intensity ~ n_recent_neighbors, data = events_h$data),
       col = "#8B2500", lwd = 2, lty = 2)
```

Each point compares, for one event, its discrete neighbour count
against its continuous Hawkes intensity. Pooled across the whole
sequence the two are only weakly correlated (Pearson's r =
`r round(cor(events_h$data$n_recent_neighbors, events_h$data$hawkes_intensity), 2)`),
and the dashed regression line is close to flat, which looks, at
first glance, like the two measures barely agree. Looking at the
shape rather than just the line tells a different story: the points
do not form one cloud but two rise-and-fall arcs, intensity climbs
with neighbour count up to a first local peak, falls back toward
zero even as the neighbour count keeps rising, then climbs to a
second, higher peak before falling again at the largest neighbour
counts. That two-arc shape lines up with the sequence's two
mainshocks. `neighbor_history()` keeps counting an event as a
"recent neighbour" for the full 3-day window regardless of how long
ago within that window it occurred, so the count stays high for a
while after a burst has already started thinning out. `hawkes_intensity()`,
with its much shorter two-day e-folding time, forgets faster: an
event late in the tail of a burst can still have a large discrete
neighbour count while its continuous intensity has already decayed
close to zero. The points with high neighbour counts but intensity
near zero, visible at both ends of the plot, are exactly these
late-in-the-burst events. Pooling two separate mainshock-aftershock
episodes into a single scatterplot is therefore what makes the
overall correlation look weak: the discrepancy is not noise, it is
the hard-window count and the fast-decaying kernel disagreeing about
how "recent" is defined, applied twice, once per mainshock.

## Venezuela: a growing catalog

Whether we go further than the map and summary above depends on
`ven_full_run` (defined earlier): with fewer than `r ven_min_n` events,
`neighbor_history()` and `hawkes_intensity()` would run without error,
but their output would not be a meaningful estimate of self-excitation,
a realistic constraint of working with live catalogs rather than a
limitation of the package. When the catalog has grown enough, the same
two-step workflow used for Amatrice is applied here as well, reusing
the same `alpha`, `beta`, and `gamma` values rather than re-tuning
them for Venezuela specifically. That reuse is deliberate for
comparability between the two sequences, but it also means the
Venezuela intensities below should be read as "what the Amatrice-scale
kernel would predict here," not as a result calibrated to this
mainshock's own magnitude or fault geometry.

```{r venezuela-full, eval = ven_full_run}
events_ven_h <- neighbor_history(events_ven, radius = 30000, lag = 3 * 86400)
events_ven_h <- hawkes_intensity(
  events_ven_h,
  alpha = 0.8, beta = 1 / (2 * 86400), gamma = 1 / 30000
)
summary(events_ven_h$data$hawkes_intensity)
plot(
  events_ven_h$data$time, events_ven_h$data$hawkes_intensity,
  type = "l", col = "#E05A2B",
  xlab = "Time (UTC)", ylab = "Hawkes intensity",
  main = "Conditional intensity: Venezuela sequence"
)
```

```{r venezuela-message, eval = !ven_full_run, echo = FALSE, results = "asis"}
cat(
  "*(Skipped: the current Venezuela catalog has", nrow(venezuela_2026),
  "events, below the", ven_min_n, "-event threshold set above.)*"
)
```

## Discussion

Applying the same functions to a complete catalog (Amatrice) and a
still-growing one (Venezuela) shows both the strength and a limit of
the workflow. `stevents` and `stgrid` handle real, irregular
spatio-temporal point patterns directly from `sf`/`POSIXct` data, and
`hawkes_intensity()` recovers a smooth, physically interpretable
excitation signal from a well-sampled sequence, visibly matching the
timing of the known Amatrice bursts and giving a graded picture that
the discrete neighbour counts alone do not. But the method is only as
good as the data feeding it, a useful reminder when applying it to
real-time monitoring rather than retrospective analysis, and it is
only as good as the parameters driving it.

Every number reported above for `alpha`, `beta`, and `gamma` was
chosen by hand, not fitted, so the intensities and the correlation
with `neighbor_history()` reflect a plausible kernel for this kind of
sequence, not a kernel estimated from the Amatrice or Venezuela data
themselves. Nothing in this vignette should be read as a claim about
the true excitation parameters of either sequence. Estimating
`alpha`, `beta`, and `gamma` from the data via maximum likelihood,
rather than setting them by hand as done here, is the natural next
step, and is exactly the estimation problem the broader thesis
addresses via an INLA-based likelihood approach.

This decomposition into background rate and self-excitation is the
core idea a Hawkes model adds on top of a simple point pattern, and
working through it here, with real aftershock data rather than a
toy example, was a useful way to build intuition for the process
before applying it elsewhere.

## LLM usage

- **LLM tool used**: Claude (Sonnet).
- **Purpose**: writing and debugging R code for the package functions and
  S3 methods, and drafting/editing the text of this vignette and the
  "Introduction to stevents" vignette.
- **Percentage of code generated by the LLM**: approximately 50-75%.
  All LLM-assisted code was reviewed, tested, and adjusted by the author
  before inclusion in the package.
