---
title: "Analysing Spatio-Temporal Wildfire Patterns with stevents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing Spatio-Temporal Wildfire Patterns with stevents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Introduction

`stevents` provides lightweight S3 classes for representing and analysing
spatio-temporal event patterns. Its primary motivation is wildfire ignition
modelling: given a set of fire events located in space and time, we want to
understand their spatial distribution, temporal intensity, and the degree to
which recent nearby events predict future ones — the self-excitation signal
at the heart of Hawkes process models.

This vignette walks through the full workflow using a simulated dataset of
wildfire ignitions on the Iberian Peninsula. All data are generated
analytically so the vignette is fully self-contained, but the spatial extent,
seasonal pattern, and clustering structure are calibrated to match historical
fire activity in the region.

## Simulating a realistic ignition dataset

The Iberian Peninsula spans roughly UTM Zone 30N easting 100,000–900,000 m
and northing 4,000,000–4,900,000 m (EPSG:25830). Historically, wildfire
activity peaks in July and August and is spatially concentrated in
north-western Iberia (Galicia, northern Portugal) and the Mediterranean
interior.

We simulate 300 ignitions over a single fire season (April–October 2023)
with two spatial clusters representing these high-risk zones, plus a
background of scattered events.

```{r simulate}
library(stevents)

set.seed(2024)

n_total     <- 300
n_cluster1  <- 120   # NW cluster: Galicia / N. Portugal
n_cluster2  <- 100   # Mediterranean interior
n_background <- 80   # scattered background

# --- Cluster 1: NW Iberia ---
x1 <- rnorm(n_cluster1, mean = 180000, sd = 40000)
y1 <- rnorm(n_cluster1, mean = 4700000, sd = 50000)

# --- Cluster 2: Mediterranean interior ---
x2 <- rnorm(n_cluster2, mean = 650000, sd = 50000)
y2 <- rnorm(n_cluster2, mean = 4450000, sd = 40000)

# --- Background: uniform over the Peninsula ---
x3 <- runif(n_background, 100000, 900000)
y3 <- runif(n_background, 4000000, 4900000)

x_all <- c(x1, x2, x3)
y_all <- c(y1, y2, y3)

# --- Temporal distribution: seasonal peak July-August ---
# We model fire season timing with a Beta distribution skewed toward
# mid-summer. Days are mapped onto April 1 - October 31 (214 days).
season_start <- as.POSIXct("2023-04-01")
season_end   <- as.POSIXct("2023-10-31")
season_days  <- as.numeric(difftime(season_end, season_start, units = "secs"))

# Beta(2, 5) gives a right-skewed distribution peaking around day 60
# (early June); Beta(5, 2) peaks late. We use Beta(3, 1.5) for a
# mid-summer peak with a long tail into autumn.
t_frac  <- rbeta(n_total, shape1 = 3, shape2 = 1.5)
t_all   <- season_start + t_frac * season_days

events <- stevents(
  x    = x_all,
  y    = y_all,
  time = t_all,
  crs  = 25830           # ETRS89 / UTM zone 30N
)
```

## The `stevents` class

### `print`

The print method gives a compact overview of the event pattern:

```{r print-stevents}
print(events)
```

### `summary`

The summary method reports derived statistics — temporal span, mean daily
rate, and spatial density:

```{r summary-stevents}
summary(events)
```

### `plot`

The plot method shows the spatial distribution of ignitions alongside the
temporal intensity curve. The seasonal peak in summer and the two spatial
clusters should be clearly visible:

```{r plot-stevents, fig.height = 4.5}
plot(events)
```

## The `stgrid` class

To aggregate events and model spatially varying intensity, we overlay a
regular space-time tessellation on the study region. We use 100 km × 100 km
spatial cells and monthly temporal bins — a resolution that balances
statistical stability with spatial detail for a season-level analysis.

```{r create-grid}
grid <- stgrid(
  bbox      = c(100000, 4000000, 900000, 4900000),
  cell_size = 100000,          # 100 km in metres
  t_window  = as.POSIXct(c("2023-04-01", "2023-10-31")),
  t_step    = "1 month",
  crs       = 25830
)
```

### `print`

```{r print-stgrid}
print(grid)
```

### `summary`

```{r summary-stgrid}
summary(grid)
```

### `plot`

A raw grid before counting has no event information; all cells are drawn
with equal fill:

```{r plot-stgrid-empty, fig.height = 5}
plot(grid)
```

## Counting events per space-time cell

`count_events()` assigns each ignition to a spatial cell and temporal bin,
returning the grid enriched with a `counts` table. This is the basic
aggregation step before fitting an intensity model.

```{r count-events}
grid_counted <- count_events(events, grid)
head(grid_counted$counts)
```

Plotting the enriched grid shades cells by total event count across the
full season. The NW and Mediterranean clusters should stand out clearly:

```{r plot-stgrid-counts, fig.height = 5}
plot(grid_counted)
```

We can also inspect a single temporal bin. Bin 2 corresponds to May, when
activity is beginning to rise:

```{r plot-stgrid-bin, fig.height = 5}
plot(grid_counted, t_bin = 2)
```

## Computing neighbour history

`neighbor_history()` computes, for each event, how many previous ignitions
occurred within a spatial radius and a temporal lag. This quantity
approximates the self-excitation signal in a Hawkes process model: a high
neighbour count suggests the event may have been triggered by recent nearby
activity rather than background conditions alone.

We use a 50 km radius and a 14-day lag — parameters consistent with the
spatial scale of fire weather persistence and the typical duration of
elevated fire risk following a large ignition event.

```{r neighbor-history}
events_hist <- neighbor_history(
  events,
  radius = 50000,      # 50 km in metres
  lag    = 14 * 86400  # 14 days in seconds
)

summary(events_hist$data$n_recent_neighbors)
```

Most ignitions have zero or very few recent neighbours, as expected for a
background-dominated process. The tail of the distribution captures the
clustered summer episodes where multiple ignitions follow one another in
quick succession within the same region.

```{r hist-neighbors, fig.height = 4}
hist(
  events_hist$data$n_recent_neighbors,
  breaks = 15,
  col    = "#FD8D3C",
  border = "#8B2500",
  main   = "Distribution of recent neighbour counts",
  xlab   = "n_recent_neighbors",
  ylab   = "Frequency"
)
```

## Summary

This vignette demonstrated the core `stevents` workflow:

1. Construct an `stevents` object from coordinates and timestamps.
2. Inspect it with `print()`, `summary()`, and `plot()`.
3. Build an `stgrid` tessellation and inspect it with the same methods.
4. Aggregate events with `count_events()` to obtain per-cell counts.
5. Enrich each event with `neighbor_history()` to capture the self-excitation
   signal relevant to Hawkes process modelling.

The resulting data structures — a classified event table with neighbour
history and a gridded count surface — are the natural inputs for a
spatio-temporal intensity model such as a log-Gaussian Cox process or a
Hawkes process fitted via INLA.
