---
title: "Analysing Spatio-Temporal Wildfire Patterns with stevents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing Spatio-Temporal Wildfire Patterns with stevents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Introduction

`stevents` provides lightweight S3 classes for representing and analysing
spatio-temporal event patterns. Its primary motivation is wildfire ignition
modelling: given a set of fire events located in space and time, we want to
understand their spatial distribution, temporal intensity, and the degree to
which recent nearby events predict future ones — the self-excitation signal
at the heart of Hawkes process models.

This vignette walks through the full workflow using a simulated dataset of
wildfire ignitions on the Iberian Peninsula. All data are generated
analytically so the vignette is fully self-contained, but the spatial extent,
seasonal pattern, and clustering structure are calibrated to match historical
fire activity in the region.

## Simulating a realistic ignition dataset

The Iberian Peninsula spans roughly UTM Zone 30N easting 100,000–900,000 m
and northing 4,000,000–4,900,000 m (EPSG:25830). Historically, wildfire
activity peaks in July and August and is spatially concentrated in
north-western Iberia (Galicia, northern Portugal) and the Mediterranean
interior.

We simulate 300 ignitions over a single fire season (April–October 2023)
with two spatial clusters representing these high-risk zones, plus a
background of scattered events. A land mask derived from Natural Earth
ensures no events fall in the sea.

```{r simulate, message = FALSE, warning = FALSE}
library(stevents)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)

set.seed(2024)

n_cluster1   <- 120
n_cluster2   <- 100
n_background <-  80
n_total      <- n_cluster1 + n_cluster2 + n_background

# --- Cluster 1: NW Iberia (Galicia / N. Portugal) ---
x1 <- rnorm(n_cluster1, mean = 180000, sd = 40000)
y1 <- rnorm(n_cluster1, mean = 4700000, sd = 50000)

# --- Cluster 2: Mediterranean interior ---
x2 <- rnorm(n_cluster2, mean = 650000, sd = 50000)
y2 <- rnorm(n_cluster2, mean = 4450000, sd = 40000)

# --- Background: uniform over the Peninsula ---
x3 <- runif(n_background, 100000, 900000)
y3 <- runif(n_background, 4000000, 4900000)

x_all <- c(x1, x2, x3)
y_all <- c(y1, y2, y3)

# --- Seasonal timing: Beta(3, 1.5) peaks in mid-summer ---
season_start <- as.POSIXct("2023-04-01")
season_end   <- as.POSIXct("2023-10-31")
season_secs  <- as.numeric(difftime(season_end, season_start, units = "secs"))
t_frac       <- rbeta(n_total, shape1 = 3, shape2 = 1.5)
t_all        <- season_start + t_frac * season_secs

# --- Land mask: crop to Peninsula, exclude Canarias and Baleares ---
iberia_raw <- ne_countries(
  country     = c("Spain", "Portugal"),
  scale       = "medium",
  returnclass = "sf"
)
peninsula_bbox <- st_bbox(
  c(xmin = -9.5, ymin = 36.0, xmax = 3.3, ymax = 44.0),
  crs = st_crs(4326)
)
iberia     <- st_crop(iberia_raw, peninsula_bbox)
iberia_utm <- st_transform(iberia, crs = 25830)

candidates <- st_as_sf(
  data.frame(x = x_all, y = y_all),
  coords = c("x", "y"),
  crs    = 25830
)
on_land <- st_within(candidates, st_union(iberia_utm), sparse = FALSE)[, 1]

events <- stevents(
  x    = x_all[on_land],
  y    = y_all[on_land],
  time = t_all[on_land],
  crs  = 25830
)
```

## The `stevents` class

### `print`

The `print` method gives a compact overview of the event pattern —
number of events, temporal extent, bounding box, and CRS:

```{r print-stevents}
print(events)
```

### `summary`

The `summary` method reports derived statistics: temporal span, mean
daily rate, and a rough measure of spatial density:

```{r summary-stevents}
summary(events)
```

### `plot`

The built-in `plot` method shows the spatial scatter alongside the
temporal intensity histogram. The seasonal peak in July–August and the
two spatial clusters are clearly visible:

```{r plot-stevents, fig.height = 4.5}
plot(events)
```

### `as_sf`: interoperability with the `sf` ecosystem

`as_sf()` converts the `stevents` object to a standard `sf` point layer,
making it immediately compatible with `ggplot2`, `tmap`, and any other
package in the spatial R ecosystem. This also allows overlaying the
events on a proper basemap of the Iberian Peninsula:

```{r map, fig.height = 5, fig.width = 7, message = FALSE}
events_sf <- as_sf(events)

ggplot() +
  geom_sf(
    data      = iberia,
    fill      = "#F5F0E8",
    colour    = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data   = st_transform(events_sf, crs = 4326),
    colour = "#E05A2B",
    alpha  = 0.5,
    size   = 1.2
  ) +
  coord_sf(xlim = c(-9.5, 3.3), ylim = c(36.0, 44.0)) +
  labs(
    title    = "Simulated wildfire ignitions — Iberian Peninsula 2023",
    subtitle = paste0(nrow(events$data), " events | land only | ETRS89 / UTM zone 30N"),
    caption  = "Simulated data for illustration purposes"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9,  colour = "grey40"),
    plot.caption  = element_text(size = 7,  colour = "grey60")
  )
```

### `subset`: filtering by space and time

`subset()` returns a new `stevents` containing only the events that
satisfy a temporal window, a spatial bounding box, or both. Here we
extract the peak fire months (July–August) and confirm that the
resulting object retains only summer events:

```{r subset-stevents}
events_summer <- subset(
  events,
  t_from = as.POSIXct("2023-07-01"),
  t_to   = as.POSIXct("2023-08-31")
)
print(events_summer)
```

The summer subset plotted on the basemap shows the same two clusters
but with higher relative density in the NW, consistent with the
historical pattern of summer fire activity in Galicia:

```{r map-summer, fig.height = 5, fig.width = 7}
events_summer_sf <- as_sf(events_summer)

ggplot() +
  geom_sf(
    data      = iberia,
    fill      = "#F5F0E8",
    colour    = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data   = st_transform(events_summer_sf, crs = 4326),
    colour = "#E05A2B",
    alpha  = 0.6,
    size   = 1.5
  ) +
  coord_sf(xlim = c(-9.5, 3.3), ylim = c(36.0, 44.0)) +
  labs(
    title    = "Peak summer ignitions — July and August 2023",
    subtitle = paste0(nrow(events_summer$data), " events"),
    caption  = "Simulated data for illustration purposes"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9,  colour = "grey40"),
    plot.caption  = element_text(size = 7,  colour = "grey60")
  )
```

## The `stgrid` class

To aggregate events and model spatially varying intensity, we overlay a
regular space-time tessellation on the study region. We use 100 km × 100 km
spatial cells and monthly temporal bins — a resolution that balances
statistical stability with spatial detail for a season-level analysis.

```{r create-grid}
grid <- stgrid(
  bbox      = c(100000, 4000000, 900000, 4900000),
  cell_size = 100000,
  t_window  = as.POSIXct(c("2023-04-01", "2023-10-31")),
  t_step    = "1 month",
  crs       = 25830
)
```

### `print`

```{r print-stgrid}
print(grid)
```

### `summary`

```{r summary-stgrid}
summary(grid)
```

### `plot`

A raw grid before counting has no event information; all cells are
drawn with equal fill:

```{r plot-stgrid-empty, fig.height = 5}
plot(grid)
```

## Counting events per space-time cell

`count_events()` assigns each ignition to a spatial cell and temporal
bin, returning the grid enriched with a `counts` table. This is the
basic aggregation step before fitting a spatially explicit intensity
model.

```{r count-events}
grid_counted <- count_events(events, grid)
head(grid_counted$counts)
```

Plotting the enriched grid shades cells by total event count across the
full season. The NW and Mediterranean clusters stand out clearly:

```{r plot-stgrid-counts, fig.height = 5}
plot(grid_counted)
```

We can also inspect a single temporal bin. Bin 4 corresponds to July,
the month of peak fire activity:

```{r plot-stgrid-bin, fig.height = 5}
plot(grid_counted, t_bin = 4)
```

## Computing neighbour history

`neighbor_history()` computes, for each event, how many previous
ignitions occurred within a spatial radius and a temporal lag. This
quantity approximates the self-excitation signal in a Hawkes process
model: a high neighbour count suggests the event may have been triggered
by recent nearby activity rather than background conditions alone.

We use a 50 km radius and a 14-day lag — parameters consistent with the
spatial scale of fire weather persistence and the typical duration of
elevated fire risk following a large ignition event.

```{r neighbor-history}
events_hist <- neighbor_history(
  events,
  radius = 50000,       # 50 km in metres
  lag    = 14 * 86400   # 14 days in seconds
)
summary(events_hist$data$n_recent_neighbors)
```

Most ignitions have zero or very few recent neighbours, as expected for
a background-dominated process. The tail of the distribution captures
the clustered summer episodes where multiple ignitions follow one
another in quick succession:

```{r hist-neighbors, fig.height = 4}
hist(
  events_hist$data$n_recent_neighbors,
  breaks = 15,
  col    = "#FD8D3C",
  border = "#8B2500",
  main   = "Distribution of recent neighbour counts",
  xlab   = "n_recent_neighbors",
  ylab   = "Frequency"
)
```

The spatial pattern of excitation can also be visualised on the
basemap. Events with at least one recent neighbour are highlighted in
dark red, revealing that self-excitation is concentrated in the
high-risk NW cluster:

```{r map-neighbors, fig.height = 5, fig.width = 7}
events_hist_sf <- as_sf(events_hist)
events_hist_sf <- st_transform(events_hist_sf, crs = 4326)
events_hist_sf$excited <- events_hist_sf$n_recent_neighbors > 0

ggplot() +
  geom_sf(
    data      = iberia,
    fill      = "#F5F0E8",
    colour    = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data  = events_hist_sf,
    aes(colour = excited),
    alpha = 0.6,
    size  = 1.2
  ) +
  scale_colour_manual(
    values = c("FALSE" = "#FDCBA8", "TRUE" = "#8B2500"),
    labels = c("No recent neighbours", "≥1 recent neighbour"),
    name   = NULL
  ) +
  coord_sf(xlim = c(-9.5, 3.3), ylim = c(36.0, 44.0)) +
  labs(
    title    = "Self-excitation signal across the Peninsula",
    subtitle = paste0(sum(events_hist_sf$excited), " of ",
                      nrow(events_hist_sf), " events with ≥1 recent neighbour"),
    caption  = "Radius = 50 km | Lag = 14 days"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9,  colour = "grey40"),
    plot.caption  = element_text(size = 7,  colour = "grey60"),
    legend.position = "bottom"
  )
```

## Hawkes process conditional intensity

The `neighbor_history()` function counts past events within a fixed
spatial radius and temporal lag — a discrete, binary approximation of
self-excitation. `hawkes_intensity()` replaces this with a continuous
exponential kernel that weights every past event by how recent and how
close it was:

$$\Lambda_i = \sum_{j:\, t_j < t_i} \alpha \cdot e^{-\beta(t_i - t_j)} \cdot e^{-\gamma \, d_{ij}}$$

The three parameters control the shape of the excitation surface:

- $\alpha$ — overall excitation magnitude
- $\beta$ — temporal decay rate (larger = faster decay)
- $\gamma$ — spatial decay rate (larger = shorter range)

We calibrate the parameters to the Iberian fire context: a temporal
decay of roughly one week and a spatial range of ~50 km, consistent
with the persistence of fire-prone meteorological conditions.

```{r hawkes-intensity}
events_hawkes <- hawkes_intensity(
  events_hist,
  alpha = 0.8,
  beta  = 1 / (7 * 86400),    # decay scale: ~1 week
  gamma = 1 / 50000            # decay scale: ~50 km
)
summary(events_hawkes$data$hawkes_intensity)
```

### Spatial distribution of Hawkes intensity

The continuous intensity can be mapped directly onto the Peninsula.
Areas of high intensity correspond to locations where multiple recent
nearby ignitions have accumulated excitation — the NW cluster is the
dominant hotspot:

```{r map-hawkes, fig.height = 5, fig.width = 7}
events_hawkes_sf <- as_sf(events_hawkes)
events_hawkes_sf <- st_transform(events_hawkes_sf, crs = 4326)

ggplot() +
  geom_sf(
    data      = iberia,
    fill      = "#F5F0E8",
    colour    = "grey60",
    linewidth = 0.3
  ) +
  geom_sf(
    data  = events_hawkes_sf,
    aes(colour = hawkes_intensity),
    size  = 1.5,
    alpha = 0.8
  ) +
  scale_colour_gradient(
    low  = "#FFF5F0",
    high = "#8B2500",
    name = "Hawkes\nintensity"
  ) +
  coord_sf(xlim = c(-9.5, 3.3), ylim = c(36.0, 44.0)) +
  labs(
    title    = "Hawkes conditional intensity — Iberian Peninsula 2023",
    subtitle = expression(paste(alpha, " = 0.8 | ",
                                beta, " ~ 1 week | ",
                                gamma, " ~ 50 km")),
    caption  = "Simulated data for illustration purposes"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9,  colour = "grey40"),
    plot.caption  = element_text(size = 7,  colour = "grey60")
  )
```

### Comparing discrete vs continuous excitation

Both `n_recent_neighbors` and `hawkes_intensity` capture the same
underlying signal but with different assumptions. The scatter plot shows
they correlate positively, but the continuous kernel captures finer
variation that the binary window misses entirely — events with zero
neighbours can still carry non-trivial Hawkes intensity from events just
outside the fixed window:

```{r compare-plot, fig.height = 4.5}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(
  events_hawkes$data$n_recent_neighbors,
  events_hawkes$data$hawkes_intensity,
  xlab = "n_recent_neighbors  (discrete)",
  ylab = "Hawkes intensity  (continuous)",
  main = "Discrete vs continuous excitation",
  pch  = 16,
  col  = adjustcolor("#E05A2B", alpha.f = 0.5),
  cex  = 0.8
)
abline(lm(hawkes_intensity ~ n_recent_neighbors,
          data = events_hawkes$data),
       col = "#8B2500", lwd = 2, lty = 2)

plot(
  events_hawkes$data$time,
  events_hawkes$data$hawkes_intensity,
  type = "l",
  col  = "#E05A2B",
  xlab = "Time",
  ylab = "Hawkes intensity",
  main = "Conditional intensity over the fire season"
)

par(old_par)
```

### Sensitivity to decay parameters

A key modelling decision is the choice of $\beta$. A fast temporal
decay concentrates excitation on very recent events; a slow decay
propagates influence further into the future:

```{r sensitivity, fig.height = 4}
betas <- c(
  "fast\n(1 day)"    = 1 / 86400,
  "medium\n(1 week)" = 1 / (7 * 86400),
  "slow\n(1 month)"  = 1 / (30 * 86400)
)

intensities <- sapply(betas, function(b) {
  h <- hawkes_intensity(events_hist, alpha = 0.8, beta = b, gamma = 1 / 50000)
  mean(h$data$hawkes_intensity)
})

barplot(
  intensities,
  col       = c("#FD8D3C", "#E05A2B", "#8B2500"),
  border    = NA,
  ylab      = "Mean Hawkes intensity",
  main      = "Sensitivity to temporal decay parameter",
  las       = 1,
  names.arg = names(betas)
)
```

The mean intensity grows with slower decay because each event's
excitation lingers longer and accumulates over the season. In practice,
$\beta$ and $\gamma$ would be estimated from the data via maximum
likelihood — a natural extension of this package toward full Hawkes
process inference with INLA.

## Summary

This vignette demonstrated the complete `stevents` workflow for
spatio-temporal wildfire analysis:

1. **Simulate** a realistic ignition dataset with seasonal and spatial
   structure, filtered to land using a Natural Earth basemap.
2. **Inspect** the `stevents` object with `print()`, `summary()`, and `plot()`.
3. **Map** events on the Iberian Peninsula via `as_sf()` and `ggplot2`.
4. **Filter** by time and space with `subset()` to isolate the peak fire season.
5. **Tessellate** the study region with `stgrid` and aggregate events
   per cell using `count_events()`.
6. **Quantify self-excitation** with `neighbor_history()` (discrete) and
   `hawkes_intensity()` (continuous exponential kernel).
7. **Visualise** the spatial distribution of excitation back on the
   basemap to identify hotspots.

The resulting enriched event table — with neighbour history and Hawkes
intensity alongside coordinates and timestamps — is the natural input
for a spatio-temporal intensity model such as a log-Gaussian Cox process
or a Hawkes process fitted via INLA.
